---
title: "TSAR Project Part 1 – Exploratory Data Analysis"
author: "Kryštof Vávra, Said Fakhri (Group 03)"
date: today
format:
  pdf:
    number-sections: true
editor: visual
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

if (!require("data.table")) install.packages("data.table")
if (!require("dplyr")) install.packages("dplyr")
if (!require("skimr")) install.packages("skimr")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("GGally")) install.packages("GGally")

library(data.table)
library(dplyr)
library(skimr)
library(ggplot2)
library(GGally)
```

```{r}
# # Random sample your group's data set from the original LCdata data set
# LCdata <- fread("LCdata.csv")
# set.seed(1)
# myLCdata <- LCdata %>% sample(5) %>% slice_sample(prop = .5)
# fwrite(myLCdata, file = "myLCdata.csv")
myLCdata <- fread("myLCdata.csv")
```

# Task 1: Business Understanding

```{r}
#| label: structure-overview

glimpse(myLCdata)
str(myLCdata)
```

## Variable Overview and Business Interpretation

The dataset `myLCdata` contains 44,369 observations and 5 variables. Based on the output of `glimpse()` and `str()`, the variables can be categorized as follows:

### Variable Description

-   **annual_inc (numeric)**\
    The self-reported annual income provided by the borrower during registration. This variable is continuous and measures repayment capacity.

-   **open_acc (integer)**\
    The number of open credit lines in the borrower's credit file. This reflects the borrower’s existing credit exposure.

-   **purpose (character)**\
    A category provided by the borrower for the loan request. (e.g., credit card refinancing, debt consolidation, small business, home improvement). This is a categorical variable.

-   **home_ownership (character)**\
    The home ownership status provided by the borrower during registration.(e.g., RENT, OWN, MORTGAGE, OTHER).

-   **inq_last_6mths (integer)**\
    The number of inquiries in past 6 months (excluding auto and mortgage inquiries). This is an indicator of recent credit-seeking behavior and potential financial stress.

------------------------------------------------------------------------

## Business Context in Peer-to-Peer Lending
<what to write here??????????>

# Task 2: Data Understanding Based on Summary Statistics

## Summary Statistics

```{r}
#| label: summary-statistics

summary(myLCdata)

# install skimr if necessary
if (!require("skimr")) install.packages("skimr")
library(skimr)

skim(myLCdata)
```

------------------------------------------------------------------------

## Interpretation and Domain Context

### 1. Missing Values (NAs)

-   Are missing values present?
-   In which variables?
-   How many?
-   Is this plausible from a business perspective?
-   Would this require data preparation?

**Discussion:** \> Write interpretation here.

------------------------------------------------------------------------

### 2. Numerical Variables

For each numeric/integer variable:

#### Variable: `__________`

-   Mean:
-   Median:
-   Minimum:
-   Maximum:
-   Standard deviation:
-   Is the distribution symmetric or skewed?
-   Are there potential outliers?
-   Do the extreme values seem realistic in the Lending Club context?

**Statistical Interpretation:** \> Write here.

**Domain Interpretation:** \> Write here.

------------------------------------------------------------------------

### 3. Categorical Variables

For each categorical variable:

#### Variable: `__________`

-   Number of categories:
-   Most frequent category:
-   Least frequent category:
-   Are categories balanced or highly unbalanced?
-   Does this distribution make business sense?

**Statistical Interpretation:** \> Write here.

**Domain Interpretation:** \> Write here.

------------------------------------------------------------------------

### 4. Key Insights from Summary Statistics

-   Most important statistical findings:
-   Potential data quality issues:
-   First hypotheses about borrower behavior:

------------------------------------------------------------------------

# Task 3: Univariate Exploratory Data Analysis (EDA) with ggplot2

## Variable 1: `__________`

### Visualization

```{r}
#| label: plot-var1

# Insert ggplot code here
```

### Why This Plot Type?

-   Why is this diagram appropriate for the variable type?
-   What can be read from this diagram type (e.g., range, distribution, frequency)?

### Interpretation

-   What does the distribution look like?
-   Are there outliers?
-   Is the distribution skewed?
-   What does this imply in the business context?
-   Does this confirm findings from Task 2?

------------------------------------------------------------------------

## Variable 2: `__________`

### Visualization

```{r}
#| label: plot-var2

# Insert ggplot code here
```

### Interpretation

> Write here.

------------------------------------------------------------------------

## Variable 3: `__________`

### Visualization

```{r}
#| label: plot-var3

# Insert ggplot code here
```

### Interpretation

> Write here.

------------------------------------------------------------------------

## Variable 4: `__________`

### Visualization

```{r}
#| label: plot-var4

# Insert ggplot code here
```

### Interpretation

> Write here.

------------------------------------------------------------------------

## Variable 5: `__________`

### Visualization

```{r}
#| label: plot-var5

# Insert ggplot code here
```

### Interpretation

> Write here.

------------------------------------------------------------------------

## Overall Observations from Univariate EDA

-   Which variables show skewness?
-   Which variables show imbalance?
-   Any surprising patterns?
-   Do visual findings confirm summary statistics?

------------------------------------------------------------------------

# Task 4: Bivariate Exploratory Data Analysis (EDA)

## Pairs Plot Overview

```{r}
#| label: ggpairs-plot

if (!require("GGally")) install.packages("GGally")
library(GGally)

ggpairs(myLCdata)
```

### Initial Observations

-   Are there visible correlations?
-   Any clear patterns?
-   Any surprising relationships?

------------------------------------------------------------------------

## Selected Insight 1

### Variable Pair: `__________` vs `__________`

### Why This Pair Is Interesting

-   What pattern was observed?
-   Why was it not obvious before?

### Stand-alone Professional Plot

```{r}
#| label: custom-plot-1

# Recreate plot using ggplot2
```

### Interpretation

-   What does this relationship indicate?
-   Business interpretation?

------------------------------------------------------------------------

## Selected Insight 2

### Variable Pair: `__________` vs `__________`

### Stand-alone Professional Plot

```{r}
#| label: custom-plot-2

# Recreate plot using ggplot2
```

### Interpretation

> Write here.

------------------------------------------------------------------------

## Adding a Third Variable

### Enhanced Plot with Third Variable (e.g., color, shape, facet)

```{r}
#| label: custom-plot-3var

# ggplot with third aesthetic mapping
```

### Why This Third Variable?

-   Why is it relevant?
-   What additional insight does it provide?
-   What was not visible before?

------------------------------------------------------------------------

## Final Bivariate Insights

-   Most important relationships discovered:
-   Implications for credit risk assessment:
-   Potential hypotheses for predictive modeling:
